---
title: "A gentle introduction to R - Deakin coding club tutorial"
author: "Simone Stevenson"
date: "31/03/2022"
output:
  pdf: default
  html: default
theme: hpstr
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

# Useful resources


- Free beginners R course: https://ashavrova.com/r-course/
- Cheat sheet for R basics: https://iqss.github.io/dss-workshops/R/Rintro/base-r-cheat-sheet.pdf 
- Cheat sheets for everything: https://www.rstudio.com/resources/cheatsheets/ 
- Ecology in R tutorials https://ourcodingclub.github.io/tutorials.html 
- Tidy data tutorials https://r4ds.had.co.nz/tidy-data.html
- Common errors and what they mean: https://ourcodingclub.github.io/tutorials/troubleshooting/ 


# Loading your data into R

To get your file path for the first time:

1. Go to 'session' tab up very top of R Studio
2. Go to 'set working directory'
3. Go to 'choose directory'
4. Navigate to the folder where you've saved your data and hit open

    It will print out to the console (bottom left panel) something like this:
    
    setwd("C:/folder1/folder2/your_project_folder")

5. Copy the quotation marks and everything in between them and add below to where
   it says: "Add directory path here", eg:

    basedir <- "C:/folder1/folder2/your_project_folder"
    setwd(basedir)
    
6. Next time you open your code you can just run the two lines above, instead of
   starting the working directory from session.


## Common errors loading data

- Error: Cannot change working directory - you have mispelled something in the file path
or you haven't created the folder yet

- Error: Cannot find file - you have mispelled the file name. Yes you have. OR, you've
saved it in the wrong place, but you've probably mispelled it.

- Error: Cannot open connection - either one of the above, or you are trying to access
a folder you don't have access to (eg if you haven't logged into the VPN and
want to use N drive)

- Error: '\U' used without hex digits in character string starting ""C:\U" - 
you've used backslashes instead of forward slashes in your file path


```{r setting directory and loading data, include=TRUE, eval=FALSE}

# Set the working directory

basedir <- "Add directory path here"

setwd(basedir)

raw_df <- read.csv("df_ful.csv") # throws an error

list.files() # 1 - checks you're looking in the right place and 2 - Gets
# the correct spelling so you can just copy and paste it

raw_df <- read.csv("df_full.csv")

View(raw_df)

```

# Using functions and packages

## Functions

- Recognise functions because they will be a word (often a verb) followed by some
brackets (eg read.csv() or mean()). Functions perform actions on data.

- Information the function needs to do it's job are called 'arguments'. These
go inside the brackets. Some arguments are essential, some of them are optional.

- To find out what arguments a function needs, you can type ? + the function name
(eg ?mean)into the console (bottom left panel) and info will pop up in the help 
panel (bottom right). Sometimes the help is really hard to understand, so you can
also just google the function until you find some information that is helpful.


## Packages

- Packages are like apps you install into R. Each one is a collection of functions
to help you do something like clean data, make plots or fit models.

- Looking for packages to do something? Try googling 'Cran task view' + whatever
you're trying to do, eg 'ecological models' or 'spatial data'. They provide
summaries of key packages for different tasks, eg:

    https://cran.r-project.org/web/views/Environmetrics.html
    
    https://cran.r-project.org/web/views/Spatial.html


```{r functions and packages, include=TRUE, eval=FALSE }

# Functions

mean(raw_df$species) # Returns NA

?mean # Look at the help documentation (or just google it)

mean(raw_df$species, na.rm = TRUE) # Add the na.rm argument to fix it

# Load packages

install.packages("tidyverse") # Only first time, use quotations!

library(tidyverse) # Every time you open your script, DON'T use quotations!


```

# Dealing with NA values

Useful data wrangling cheat sheet
https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf

```{r remove NA, include=TRUE, eval=FALSE}

# Which columns have NA values and how many?

colSums(is.na(raw_df))

# Using some functions from the tidyverse package, lets see how the NAs are 
# distributed among groups

raw_df %>% # This symbol is the 'pipe', essentially means 'and then'
  group_by(taxon) %>% # Do this - group by
  summarise(NA_count = sum(is.na(human_background_type)),
            NA_proportion = NA_count/n())

# Remove all rows without complete data for every column

df_complete <- raw_df %>% 
               na.omit()

# Check the size of your complete data vs your raw data
nrow(raw_df)
nrow(df_complete)

```

# Saving data

```{r saving, include=TRUE, eval=FALSE}

# The first argument is the object you want to save, the second argument is what you want to name the file
write.csv(df_complete, "df_NA_removed.csv")

```

# Simple plots

For a visual guide where you can select the plot then get the code for it,
visit:

https://r-graph-gallery.com/

Useful base R graphics cheat sheet

http://publish.illinois.edu/johnrgallagher/files/2015/10/BaseGraphicsCheatsheet.pdf

```{r plot full dataset, include=TRUE, eval=FALSE}

hist(raw_df$comments, 50)

table(raw_df$taxon)

plot(raw_df$followers, raw_df$likes)

boxplot(likes ~ human_presence, data = raw_df, frame = FALSE)

# We can check stack overflow for how to remove outliers from the plot
# https://stackoverflow.com/questions/14207739/how-to-remove-outliers-in-boxplot-in-r


# 1. Open tiff file
tiff("220401_boxplot.tiff", units="in", width=10, height=5, res=200)

# 2. Create the plot

plot <- boxplot(likes ~ human_presence, data = raw_df, frame = FALSE,outline=FALSE) # Add outline argument to remove outliers

# 3. Close the file

dev.off()

```









