---
title: "Leaflet tutorial"
author: "Simone Stevenson"
date: "19/07/2022"
output: html_document
self_contained: yes
---


```{r setup, include=FALSE, warning=FALSE, error=FALSE, message=FALSE}

# https://bookdown.org/yihui/rmarkdown/r-code.html

knitr::opts_chunk$set(echo = FALSE, message = FALSE,
  warnings = FALSE, error = FALSE)

```

# Libraries

```{r libraries, warning=FALSE, error=FALSE, message=FALSE}

if (!require(pacman)) install.packages('pacman')

# Install packages as needed (after installing pacman, it will install any additional
# packages you don't already have)


pacman::p_load(tidyverse, #data wrangling
               leaflet, leaflet.extras, # leaflet package  + extras
               sf, # For handling polygons
               rnaturalearth,  
               rnaturalearthdata, # For downloading country shapefiles
               spData, # For some data we are ging to use 
               htmlwidgets,# to save maps
               ggmap # for automatically pulling coords for places you know (i.e. to centre your map)
               )

```

# Load data

Let's use some data from the spData package 

## lnd

spData: Polygons representing large administrative zones in London

• NAME Borough name
• GSS_CODE Official code
• HECTARES How many hectares
• NONLD_AREA Area outside London
• ONS_INNER Office for national statistics code
• SUB_2009 Empty column
• SUB_2006 Empty column
• geometry sfc_MULTIPOLYGON


## cycle_hire

spData: Points representing cycle hire points across London

FORMAT:
• id Id of the hire point
• name Name of the point
• area Area they are in
• nbikes The number of bikes currently parked there
• nempty The number of empty places
• geometry sfc_POINT

## cycle_hire_osm

spData: Dataset downloaded using the osmdata package representing cycle hire points across London

• osm_id The OSM ID
• name The name of the cycle point
• capacity How many bikes it can take
• cyclestreets_id The ID linked to cyclestreets’ photomap
• description Additional description of points
• geometry sfc_POINT

## uk

rnaturalearthdata: Polygon of the United Kingdom's boundary

```{r data, warning=FALSE, error=FALSE, message=FALSE, echo = FALSE}

# spData: Polygons representing large administrative zones in London
data(lnd)
summary(lnd)

# spData: Points representing cycle hire points across London
data(cycle_hire)
summary(cycle_hire)

# spData: Dataset downloaded using the osmdata package representing cycle hire points across London
data(cycle_hire_osm)
summary(cycle_hire_osm)


# rnaturalearthdata: Polygon of the United Kingdom's boundary
uk <- ne_countries(scale = "medium", returnclass = "sf", country = "United Kingdom")

# re-project so this boundary matches our others
# uk <- st_transform(uk, crs = st_crs(lnd))

```

#Static plot

```{r static plot, warning=FALSE, error=FALSE, message=FALSE}

plot(st_geometry(uk))
plot(st_geometry(lnd), add = TRUE)
plot(st_geometry(cycle_hire), add = TRUE, col = "deeppink")
plot(st_geometry(cycle_hire_osm), add = TRUE, col = "turquoise")

```
# Combine cycle data

You could leave them and add as separate layers but this will make things a bit simpler

```{r combine, warning=FALSE, error=FALSE, message=FALSE, echo = FALSE}

cycle_hire_osm_temp <- cycle_hire_osm %>% 
                       mutate(id = paste0("osm", osm_id)) %>% 
                       rename(nbikes = capacity) %>% 
                       dplyr::select(id, name, nbikes, geometry)

cycle_hire_temp <- cycle_hire  %>% 
                   dplyr::select(id, name, nbikes, geometry)

cycle_hire_all <- rbind(cycle_hire_osm_temp, cycle_hire_temp) 

rm(cycle_hire_temp, cycle_hire_osm_temp)

```


# Add points and polygons

You can zoom in manually but it's not ideal

```{r basic map, warning=FALSE, error=FALSE, message=FALSE}

#initialise the leaflet map
leaflet() %>% # Use the pipe operator to add more layers to map
  # Add a base map background - this one is the default that comes with leaflet
     addTiles(group = "Basemap") %>%
  # Add the polygon of the uk
     addPolygons(data = uk) %>% 
  # Add our polygon data: lnd (london borough boundaries)
     addPolygons(data = lnd) %>% 
  # Add our points (cycle hire stations) and colour them pink
    addCircles(data = cycle_hire_all, color = "deeppink") 

```

# Centre the map

To obtain a google map API key so you can use the ggmap library and geocode function, go to https://mapsplatform.google.com/ - it's really easy

```{r centre, warning=FALSE, error=FALSE, message=FALSE}

# To register your API use the commented out code below:
# ggmap::register_google(key = "mQkzTpiaLYjPqXQBotesgif3EfGL2dbrNVOrogg", write = TRUE) # this is a fake key

# Use geocode (OR just get the coords from google maps!)
mapCentre <- geocode("London") # Use an address or landmark


leaflet() %>% 
     addTiles(group = "Basemap") %>% 
     addPolygons(data = uk) %>% 
     addPolygons(data = lnd) %>% 
 addCircles(data = cycle_hire_all, color = "deeppink") %>% 
# Set the map centre with lat lon coordinates, set default zoom in level
 setView(lng = mapCentre$lon, lat = mapCentre$lat, zoom = 10)

```


# Change the basemap

- print the word 'providers' into the console (no quotation marks) to get a list of background names,

- Or, get previews of what they look like at: https://leaflet-extras.github.io/leaflet-providers/preview/

```{r basemaps, warning=FALSE, error=FALSE, message=FALSE}

# you can print the word providers to see all of the option names
leaflet() %>% 
  # Use addProviderTiles to select a different basemap
  # Can either put just the name in quotation marks "OpenStreetMap.Mapnik" or use the format providers$name - the latter is better if you are going to share the map with others
     addProviderTiles(providers$Esri.WorldGrayCanvas) %>% 
     addPolygons(data = lnd) %>% 
     addCircles(data = cycle_hire_all, color = "deeppink") %>% 
     setView(lng = mapCentre$lon, lat = mapCentre$lat,zoom = 10)

```

# Colour data by a variable of interest

```{r colour points, warning=FALSE, error=FALSE, message=FALSE}

# Create a colour palette for the variable you want colour to vary by
# If its numeric/continuous use colorNumeric
# This will colour our cycle hire stations by the number of bikes they hold
cycleHirePal <- colorNumeric("inferno", domain = cycle_hire_all$nbikes, reverse = TRUE)

# If it is a factor use colorFactor
# This will colour the different boroughs
lndPal <- colorFactor("viridis", domain = as.factor(lnd$NAME))

# Colour the polygons by borough and points by number of bikes available

leaflet() %>% 
     addProviderTiles(providers$Esri.WorldGrayCanvas) %>%  
     # Add the fill colour and opacity arguments to colour the polygons, using the colour palette you made above
     addPolygons(data = lnd, group = "NAME",
              fillColor = ~lndPal(NAME),
              fillOpacity = 0.8) %>% 
  # For the circles use color not fill, same as ggplot
     addCircles(data = cycle_hire_all, color = ~cycleHirePal(nbikes)) %>% 
     setView(lng = mapCentre$lon, lat = mapCentre$lat,zoom = 10)

# Make the borders nicer

leaflet() %>% 
     addProviderTiles(providers$Esri.WorldGrayCanvas) %>% 
     addPolygons(data = lnd, group = "NAME",
              fillColor = ~lndPal(NAME), 
              fillOpacity = 0.8,
              #Add the weight argument to make the borders thinner
              weight = 1) %>%
     addCircles(data = cycle_hire_all, color = ~cycleHirePal(nbikes)) %>%
     setView(lng = mapCentre$lon, lat = mapCentre$lat,zoom = 10)

# Change the colour of the border when you hover over it

leaflet() %>% 
     addProviderTiles(providers$Esri.WorldGrayCanvas) %>% 
     addPolygons(data = lnd, group = "NAME",
              fillColor = ~lndPal(NAME),
              fillOpacity = 0.8,
              # Add the color argument to change the border colour
              color = "white",
              weight = 1,
              # Use highlight options to change the apparance of a polygon if you hover your mouse
              highlightOptions =
                # color = border colour will change when you click
                # weight = make it thicker
                # bringToFront = bring the polygons to the front when you hover
                highlightOptions(color='deeppink',weight=2,
                                                  bringToFront = TRUE)) %>% 
     addCircles(data = cycle_hire, color = ~cycleHirePal(nbikes)) %>%
     setView(lng = mapCentre$lon, lat = mapCentre$lat,zoom = 10) 

```

# Add popup labels


```{r popups, warning=FALSE, error=FALSE, message=FALSE}

leaflet() %>% 
     addProviderTiles(providers$Esri.WorldGrayCanvas) %>% 
     addPolygons(data = lnd, group = "NAME",
              fillColor = ~lndPal(NAME),
              fillOpacity = 0.8,
              color = "white",
              weight = 1,
              highlightOptions = highlightOptions(color='deeppink',weight=2,
                                                  bringToFront = TRUE),
              # Add the popup argument to make more information pop up when you click on the polygon. You can paste together different values from the dataframe. The "<b>" type stuff is html code to format the text (bold, new line etc)
              popup = ~ paste("<b>",NAME, "</b>", "<br/>", HECTARES, "hectares", sep = " ")) %>%
     addCircles(data = cycle_hire_all, color = ~cycleHirePal(nbikes)) %>%
     setView(lng = mapCentre$lon, lat = mapCentre$lat,zoom = 10)



leaflet() %>% 
     addProviderTiles(providers$Esri.WorldGrayCanvas) %>%  
     addPolygons(data = lnd, group = "NAME",
              fillColor = ~lndPal(NAME),
              fillOpacity = 0.8,
              color = "white",
              weight = 1,
              highlightOptions = highlightOptions(color='deeppink',weight=2),
              popup = ~ paste("<b>",NAME, "</b>", "<br/>", HECTARES, "hectares", sep = " ")) %>% 
  # Add popups to our points as well
     addCircles(data = cycle_hire_all, color = ~cycleHirePal(nbikes),
                      popup = ~ paste("<b>",name, "</b>", "<br/>", nbikes, "bikes", sep = " ")) %>%
     setView(lng = mapCentre$lon, lat = mapCentre$lat,zoom = 10)

```

# Which borough has the most bike stations, vs which has the most bikes, per hectare?


```{r capacity of boroughs, warning=FALSE, error=FALSE, message=FALSE}

# Intersect points with boroughs, to assign a borough name to each hire station
tmp <- st_intersection(cycle_hire_all, lnd) %>% 
  # For each borough (represented by name)
       group_by(NAME) %>% 
  # Calculate the number of bikes, stations, and density of these things per hectare
       mutate(numberOfBikes = sum(nbikes, na.rm = TRUE),
              numberOfHireStations = n_distinct(id),
              bikeDensity = numberOfBikes/HECTARES,
              hireStationDensity = numberOfHireStations/HECTARES) %>% 
       #Remove the spatial info from this dataframe because we will merge it with one of the others anyway
       st_drop_geometry() %>%
       # Select just the borough name and the new values we have calculated
       dplyr::select(NAME, numberOfHireStations, hireStationDensity,
                     numberOfBikes, bikeDensity) %>% 
      # Ungroup so any further calculations aren't applied to the groups
       ungroup() %>% 
      # Remove duplicates
       distinct()

# Merge the new values into our london polygons
lnd2 <- lnd %>% 
        merge(tmp, by = "NAME", all = TRUE) %>% 
        # Convert NA to 0
        mutate(hireStationDensity = ifelse(is.na(hireStationDensity), 0, hireStationDensity))

# Remove the tmp data
rm(tmp)

# Make a colour palette for the station hire density variable
stationDensityPal <- colorNumeric("inferno", domain = lnd2$hireStationDensity, 
                        alpha = FALSE, reverse = TRUE)

# And one for the bike density per boroughs
bikeDensityPal <- colorNumeric("inferno", domain = lnd2$bikeDensity, 
                        alpha = FALSE, reverse = TRUE)

```

```{r compare}

# Make another palette for the number of bikes 

bikeDensityPal <- colorNumeric("magma", domain = lnd2$bikeDensity, 
                        alpha = FALSE, reverse = TRUE)

# Make a map that shows the density of bikes in the boroughs - am I likely to find a bike at a station where I am?

leaflet() %>% 
     addProviderTiles(providers$Esri.WorldGrayCanvas) %>% 
     addPolygons(data = lnd2, group = "NAME",
              fillColor = ~bikeDensityPal(bikeDensity),
              fillOpacity = 0.8,
              color = "white",
              weight = 1,
              highlightOptions = highlightOptions(color='deeppink',weight=2),
              popup = ~ paste("<b>",NAME, "</b>", "<br/>", 
                              round(bikeDensity,3), "bikes per hectare", 
                              sep = " ")) %>% 
     setView(lng = mapCentre$lon, lat = mapCentre$lat,zoom = 10) %>% 
  # Add a legend to match our palette
     addLegend(data = lnd2, "topright", pal = bikeDensityPal, values = ~ bikeDensity,
    title = "Bikes per hectare",
    opacity = 1
  )

# Make the same map with station density so we can compare (how far do I have to walk to find a station?)

leaflet() %>% 
     addProviderTiles(providers$Esri.WorldGrayCanvas) %>% 
     addPolygons(data = lnd2, group = "NAME",
              fillColor = ~stationDensityPal(hireStationDensity),
              fillOpacity = 0.8,
              color = "white",
              weight = 1,
              highlightOptions = highlightOptions(color='deeppink',weight=2),
              popup = ~ paste("<b>",NAME, "</b>", "<br/>", 
                              round(hireStationDensity,3), "hire stations per hectare", 
                              sep = " ")) %>%
       addLegend(data = lnd2, "topright", pal = stationDensityPal, values = ~ hireStationDensity,
        title = "Hire stations per hectare",
        opacity = 1
      )
```

# Add the hire stations back in

```{r bike capacity map, warning=FALSE, error=FALSE, message=FALSE}

leaflet() %>%
     addProviderTiles(providers$Esri.WorldGrayCanvas) %>%
     addPolygons(data = lnd2, group = "NAME",
              fillColor = ~bikeDensityPal(bikeDensity),
              fillOpacity = 0.8,
              color = "white",
              weight = 1,
              highlightOptions = highlightOptions(color='deeppink',weight=2),
              popup = ~ paste("<b>",NAME, "</b>", "<br/>", 
                              round(bikeDensity,3), "bikes per hectare", 
                              sep = " ")) %>% 
    #Use addCircleMakers instead of addCircles so we can change size
    addCircleMarkers(data = cycle_hire_all,
    # Add the arument 'radius' and the variable you want to size by.
    # I think sqrt normalises your data
    radius = ~ sqrt(nbikes),
    stroke = FALSE, fillOpacity = 0.5,
    fillColor = "white",
    popup = ~ paste("<b>",name, "</b>", "<br/>", nbikes, "bikes", sep = " ")) %>% 
    addLegend(data = lnd2, "topright", pal = bikeDensityPal, values = ~ bikeDensity,
    title = "Bikes per hectare",
    opacity = 1) %>% 
    setView(lng = mapCentre$lon, lat = mapCentre$lat,zoom = 12)  

```

# Toggle multiple data layers

```{r toggle data, warning=FALSE, error=FALSE, message=FALSE}

leaflet() %>% 
     addProviderTiles(providers$Esri.WorldGrayCanvas) %>% 
     addPolygons(data = lnd2, group = "NAME",
              fillColor = ~bikeDensityPal(bikeDensity),
              fillOpacity = 0.8,
              color = "white",
              weight = 1,
              highlightOptions = highlightOptions(color='deeppink',weight=2),
              popup = ~ paste("<b>",NAME, "</b>", "<br/>", 
                              round(bikeDensity,3), "bikes per hectare", 
                              sep = " ")) %>% 
    addCircleMarkers(data = cycle_hire_all,
    radius = ~ sqrt(nbikes),
    stroke = FALSE, fillOpacity = 0.5,
    fillColor = "white",
    popup = ~ paste("<b>",name, "</b>", "<br/>", nbikes, "bikes", sep = " "),
    # Add a 'group' argument with the name you want your layer ot have when it appears on the map
    group = "number of bikes at station") %>%
   # Use addLayersControl to add the toggle on and off legend. Names in overlayGroups need to match the names you supplied to 'group' in the layer
    addLayersControl(overlayGroups = c("bikes per hectare", "number of bikes at station"),
                      position = "bottomright",
                     # I can't remember what this argument does lol
                      options = layersControlOptions(collapsed = FALSE))  %>%
    addLegend(data = lnd2, "topright", pal = bikeDensityPal, values = ~ bikeDensity,
    title = "Bikes per hectare",
    opacity = 1) %>% 
    setView(lng = mapCentre$lon, lat = mapCentre$lat,zoom = 11)

```



# Toggle a single data layer

```{r split data, warning=FALSE, error=FALSE, message=FALSE}

# What if you want to toggle values of a single layer? You need to add them as separate layers to the map

# Select all hire stations where number of bikes is unknown/NA
hire_points_na <- cycle_hire_all %>%
                 filter(is.na(nbikes))

# Select ones where it is known
hire_points <- cycle_hire_all %>%
                 filter(!is.na(nbikes))


bikeMap <- leaflet() %>% 
     addProviderTiles(providers$Esri.WorldGrayCanvas) %>% 
     addPolygons(data = lnd2, group = "NAME",
              fillColor = ~bikeDensityPal(bikeDensity),
              fillOpacity = 0.8,
              color = "white",
              weight = 1,
              highlightOptions = highlightOptions(color='deeppink',weight=2),
              popup = ~ paste("<b>",NAME, "</b>", "<br/>", 
                              round(bikeDensity,3), "bikes per hectare", 
                              sep = " ")) %>% 
  # Add points where nbikes is known
    addCircleMarkers(data = hire_points,
    radius = ~ sqrt(nbikes),
    stroke = FALSE, fillOpacity = 0.5,
    fillColor = "white",
    popup = ~ paste("<b>",name, "</b>", "<br/>", nbikes, "bikes", sep = " "),
    group = "number of bikes at station") %>%
    addLayersControl(overlayGroups = c("bikes per hectare", "number of bikes at station"),
                      position = "bottomright",
                      options = layersControlOptions(collapsed = FALSE))  %>%
  # Add the NA values separately
    addCircleMarkers(data = hire_points_na,
    radius = ~ sqrt(nbikes),
    stroke = FALSE, fillOpacity = 0.5,
    fillColor = "black",
    popup = ~ paste("<b>",name, "</b>", "<br/>", nbikes, "bikes", sep = " "),
    group = "stations without bikes") %>%
  # Make sure you add each layer to you layer control tool
    addLayersControl(overlayGroups = c("bikes per hectare", "number of bikes at station", "stations without bikes"),
                      position = "bottomright",
                      options = layersControlOptions(collapsed = FALSE))  %>%
    addLegend(data = lnd2, "topright", pal = bikeDensityPal, values = ~ bikeDensity,
    title = "Bikes per hectare",
    opacity = 1) %>% 
    setView(lng = mapCentre$lon, lat = mapCentre$lat,zoom = 11)

bikeMap

```


# Find a bike station

```{r clusters, warning=FALSE, error=FALSE, message=FALSE}

# Maybe we actually need to find a bike station and the map above isn't very useful

clusterMap <- leaflet() %>% 
  # Add a basemap that includes street names
     addProviderTiles(providers$OpenStreetMap.Mapnik) %>% 
     addPolygons(data = lnd2, group = "NAME",
              fillColor = ~stationDensityPal(hireStationDensity),
              fillOpacity = 0.4,
              color = "white",
              weight = 1,
              highlightOptions = highlightOptions(color='deeppink',weight=2,
                                                  bringToFront = TRUE),
              popup = ~ paste("<b>",NAME, "</b>", "<br/>", 
                              round(hireStationDensity,3), "bikes per hectare", 
                              sep = " ")) %>% 
     addCircleMarkers(data = hire_points, color = ~cycleHirePal(nbikes),
                popup = ~ paste("<b>",name, "</b>", "<br/>", nbikes, "bikes", 
                                sep = " "),
                group = "number of bikes for hire",
                # Add this cluster options argument to cluster points together until you zoom in
                clusterOptions = markerClusterOptions()) %>%
     addCircleMarkers(data = hire_points_na, color = "darkgrey",
                popup = ~ paste("<b>",name, "</b>", "<br/>", nbikes, "bikes", 
                                sep = " "),
                group = "capacity unknown",
                clusterOptions = markerClusterOptions()) %>%
     setView(lng = mapCentre$lon, lat = mapCentre$lat,zoom = 10) %>% 
     addLayersControl(overlayGroups = c("bikes per hectare", "number of bikes for hire", 
                                        "capacity unknown"),
                      position = "bottomright",
                      options = layersControlOptions(collapsed = FALSE))  %>% 
  # If you want a layer to be clicked off as the default, hide it like this, then the user has ot choose to click it on
     hideGroup(c("capacity unknown"))

clusterMap

```

# Save maps


```{r save, warning=FALSE, error=FALSE, message=FALSE}

saveWidget(bikeMap, file=paste0( getwd(), "/londonBikeMap.html"))

saveWidget(clusterMap, file=paste0( getwd(), "/londonCyclingClusterMap.html"))

```

