---
title: "Functions tutorial"
author: "Simone Stevenson"
date: "19/07/2022"
output:
  pdf_document: default
  html_document: default
self_contained: yes
---

# Benefits of writing functions

- readability
- less typing/effort
- fewer errors
- reuse code across projects




# Libraries

```{r libraries, warning=FALSE, error=FALSE, message=FALSE}

library(tidyverse) # wrangle data
library(gridExtra) # display multiple plots
library(pokemon) # loads pokemon data
library(kableExtra) # makes nice tables

```



```{r code that could be a function}

# See how these three chunks of code are identical except for the species name in filter?
setosa_summary1 <- iris %>% 
               filter(Species == "setosa") %>% 
               summarise(mean_sepal_length = mean(Sepal.Length),
                         mean_sepal_width = mean(Sepal.Width),
                         mean_petal_length = mean(Petal.Length),
                         mean_petal_width = mean(Petal.Width))

versicolor_summary1 <- iris %>% 
                   filter(Species == "versicolor") %>% 
                   summarise(mean_sepal_length = mean(Sepal.Length),
                             mean_sepal_width = mean(Sepal.Width),
                             mean_petal_length = mean(Petal.Length),
                             mean_petal_width = mean(Petal.Width))

virginica_summary1 <- iris %>% 
                  filter(Species == "virginica") %>% 
                  summarise(mean_sepal_length = mean(Sepal.Length),
                             mean_sepal_width = mean(Sepal.Width),
                             mean_petal_length = mean(Petal.Length),
                             mean_petal_width = mean(Petal.Width))

setosa_summary1

```

```{r convert to function}

# We can write a function to cut down the repeating code.

get_species_summary <- function(name) {
  
  iris %>% 
  filter(Species == name) %>% 
  summarise(mean_sepal_length = mean(Sepal.Length),
                             mean_sepal_width = mean(Sepal.Width),
                             mean_petal_length = mean(Petal.Length),
                             mean_petal_width = mean(Petal.Width),
            Species = name)
}

setosa_summary <- get_species_summary("setosa")

versicolor_summary <- get_species_summary("versicolor")

virginica_summary <- get_species_summary("virginia")

# And if we want to change something, like add a group column, we only need
# to change our code in one place instead of 3

setosa_summary

```
# Tips and tricks
## Naming functions

- Variable or objects are nouns (eg df), functions are usually verbs (eg filter)

Best practice to include verbs in a function - can you think of examples that don't
follow this practice?

Understanding code is more important than quick typing, so pick a coherent name 
for your function

## Organising functions in your code

- put em up top
- source
- R markdown - need to go in their own chunk

# Outputs

Functions return the final value so you can assign outside

# Naming arguments

Be very general - use 'df' not 'yarra_site_1', for example
Try not to use names that are not already function names (eg mean, sample)

# Write a function template

# Syntax for writing a function

my_fun <- function(arg1, arg2) {

# Do something with args 1 and 2

}

This bit is called the function's 'signature': function(arg1, arg2)
Can think of it as the user interface. The parts inside are the 'body'

```{r template}

# Making a function

add_ten <- function(x) { #open brackets
  
  x + 10 # do something in the body
  
} #close brackets

# Using a function

add_ten(10)

add_ten(100)

```


# Arguments

Arguments are the bits of code that aren't duplicated across repeating code (for example
different vectors you want to take the mean of, different dataframes for 
different sampling sites, or species names like in the first example). These go inside
the parentheses in the function's signature.

Number of arguments technically unlimited but more == difficult to debug

## Let's start by examining arguments of functions we already know

How do we know what arguments a function takes? Let's look at the function 'cor'


```{r gets args}

args(cor)
formals(cor)
?cor # more detail
cor() #hover ur mouse

```


What types or arguments are there?

- Data arguments (what are we going to perform actions on?)
- Detail arguments (how are we going to do it?)

```{r}

# Let's look at cor() again, which is which?
?cor

# What about sample?

?sample

```


How do you enter the arguments to a function?

1. Enter them by position order (look up the order using args(function))
2. Enter them by name (name = value)
3. Common/mandatory arguments by position, rare/optional arguments by name (best practice/most 
common convention)


```{r entering arguments}

args(median)

# Entering arguments - what do we think of this?

median(TRUE, x = iris$Sepal.Length)

# How else might we do it?
median(iris$Sepal.Length, na.rm = TRUE)

```

# Fill in an argument in the body

```{r arguments}

multiply_things <- function(x, multiplier) { #open brackets
  
  x * multiplier # what goes here?
  
  
}

multiply_things(10,5)

```

# Fill in the arguments in the signature

```{r arg signature}

subtract_things <- function(x,y){
  
x - y
  
}

```

# Add a default value for one of the arguments

```{r defaults}

add_things <- function(x, n){
  
x + n
  
}

# Use the function
add_things(10, 5)

# What if we forget an argument?

add_things(10)

# What if we want to put in a default value to add, when the user doesn't specify it?

add_things_fancy <- function(x, n = 5){
  
x + n
  
}

add_things_fancy(10)

```
# Add your own default value for replace argument

```{r add ur own default}

replace_na <- function(x, replace = 0) {
  
  ifelse(is.na(x), replace, x) 
  
}

replace_na(NA,0)

replace_na(NA)

```

# Add documentation

```{r document}

#' Replace NA Values in a Vector
#'
#' This function replaces NA values in a vector with a specified replacement value.
#'
#' @param x A vector containing elements with possible NA values.
#' @param replace The value to replace NA values with.
#' @return A new vector with NA values replaced by the specified replacement value.
#'
#' @examples
#' x <- c(1, NA, 3, NA, 5)
#' result <- replace_na(x, 0)
#' result
#'
#' @export

replace_na <- function(x, replace) {
  
  # Continue with function
  ifelse(is.na(x), replace, x) 
  
}

replace_na(NA, "wrong")

```

# Add a check to the input arguments 

This is called an assertion, there are lots of shortcut packages to make writing
these easier

The stop function abandons the function and returns an error message (this is where
all those annoying error messages in your console are born)

```{r check}

add_things <- function(x, n = 5){

if(!is.numeric(x)) {stop("wrong")}
  
x + n
  
}

add_things("green")

```

# Add a numeric check to our replace_na function

```{r na check}

replace_na <- function(x, replace) {
  
  # Check that replace is numeric
  
  
  # Continue with function
  ifelse(is.na(x), replace, x) 
  
}

replace_na(NA, "wrong")

```


# Write the body of this function

```{r write body}

divide_things <- function(x, n){
  
 x/n
  
}

divide_things(9,3)

```

# Write a full function - signature and body

Write a function to take the mean of a vector, then add a value to it

Add the roxygen documentation as well

```{r write function}

add_to_mean <- function(x,y, ...) {
  
  mean(x, ...) + y
  
}

add_to_mean(c(1,2,3,NA), 100, na.rm = TRUE)

```

# Make some plots

Functions don't just take individual values, or return numbers. 

Perhaps you want to make an identical plot for each of your sampling sites or species?

Here is a simple function to make a histogram of the sepal length column in
the iris dataframe

```{r histo}

summary(iris) # Have a quick look at our df

plot_sepal_length_hist  <- function(df) {
 
   ggplot(df, # The dataframe our column is in
          aes(x = df$Sepal.Length)) + # Look at the distribution of values in the sepal length column
      geom_histogram(color = "black",
          fill = "lightblue") # We want our plot to be a histogram
  
}

plot_sepal_length_hist(iris)

```

Add an argument to the function that allows us to change the colour of the bars

BONUS CHALLENGE, you could also:

Add an argument to the function that lets us change how many bins to use

Add an argument to the function that lets us add a title

```{r histo colors}

plot_sepal_length_hist  <- function(df, colour = "lightblue") {
 
   ggplot(df, # The dataframe our column is in
          aes(x = df$Sepal.Length)) + # Look at the distribution of values in the sepal length column
      geom_histogram(color = "black",
          fill = colour ) # We want our plot to be a histogram
  
}

plot_sepal_length_hist(iris, "turquoise")

```
# How to specify column names as arguments

What if we don't want to plot Sepal Length? Maybe we want to plot Sepal Width

Using double curly braces allows us to use a lot of tidyverse functions that
require you to specify a column name without quotation marks

This can make a function a lot more widely applicable in some contexts

```{r more general histogram function}

# Function to plot histograms for any variable
plot_histogram <- function(df, x_variable, col) {
  
  # Because ggplot (and other tidyverse functions) don't use quotation marks,
  # we need to wrap our argument that specifies a column name in double curly braces
  
  ggplot(df, aes(x = {{ x_variable }})) + 
    geom_histogram(fill = col, color = "black") +
    labs(
      x = deparse(substitute(x_variable)),
      y = "Frequency"
    )
}

# Now we can apply the function to completely different columns 

sepal_width <- plot_histogram(iris, Sepal.Width, "pink")

petal_length <- plot_histogram(iris, Petal.Length, "lightblue")

grid.arrange(sepal_width, petal_length, ncol = 2)

# Or even datasets

plot_histogram(mtcars, mpg, "green4")

```

Here's another example for selecting columns

```{r select a column}

get_max <- function(df, column_name){
  
  df %>% 
    select({{column_name}}) %>% 
    filter({{column_name}} == max({{column_name}})) %>% 
    pull(.)
  
}

get_max(iris, Sepal.Length)

get_max(mtcars, mpg)

```

# Have a go at using the double curly braces to finish this function

```{r curly try}

add_ten_to_column <- function(df, column_name) {
  
  df %>% 
    mutate(new_column = ____ + 10)

}

new_iris <- add_ten_to_column(iris, ____ )

head(new_iris)

```

# Convert code into a function

You can follow these steps to convert your existing code into a function:

1. Create the template with function name, signature and body
2. Paste the script into the body
3. Choose which parts of your script should be the arguments
4. Replace specific values with argument names
5. Make specific variable names more general
6. Remove the final assignment of your output to an object

There are also shortcuts I will show you a bit later :)

Turn this code into a function that just takes a dataset, and a character value
from the column egg_group_1 as its arguments.

```{r make this code into a function}

cheat_pokemon_function <- function(pokemon, egg_group_1) {
  # Filter the dataset for "monster" in egg_group_1
  monster_pokemon <- pokemon %>%
    filter(egg_group_1 == "monster")
  
  summary_table <- monster_pokemon %>% 
                   summarise(egg_group_1 = "monster",
                             species_count = n_distinct(pokemon),
                             mean_height = mean(height, na.rm = TRUE),
                             max_speed = max(speed, na.rm = TRUE))
  
  # Print the table
  summary_table
}

```


```{r pokemon function}

summarise_pokemon <- function(x, egg_group_name){
  
  # Filter the dataset for "monster" in egg_group_1
monster_pokemon <- x %>%
  filter(egg_group_1 == egg_group_name)

monster_pokemon %>% 
                 summarise(egg_group_1 = egg_group_name,
                           species_count = n_distinct(pokemon),
                           mean_height = mean(height, na.rm = TRUE),
                           max_speed = max(speed, na.rm = TRUE))


}

summarise_pokemon(pokemon, "monster")


summarise_pokemon(pokemon, "bug")

```

# Now add roxygen documentation

```{r pokemon documentation}

```

# Extra challenge 1 - select columns

Make the same function, except in this one, you can select different columns
to take the mean and speed of, then name them accordingly

```{r fancy summary}

fancy_summary <- function(df, group, mean_var, max_var, mean_name, max_name){
  
  group_data <- ____ %>% 
                filter(egg_group_1 == ____ )
  
  summary_table <- group_data %>%
                   summarise(egg_group_1 = ____,
                             species_count = n_distinct(pokemon),
                             ____ ,
                             ____ )

  names(summary_table) <- c("egg_group_1", "species_count", ___ , ___ )

  # Print the table
  summary_table
  
}

fancy_summary()

```


# Extra challenge 2 - use a loop

```{r loop}

# Split data by species

setosa_data <- iris %>% 
               filter(Species == "setosa") 

versicolor_data <- iris %>% 
                   filter(Species == "versicolor")

virginica_data <- iris %>% 
                  filter(Species == "virginica") 


iris_species_list <- list(setosa_data, versicolor_data, virginica_data)
species_colors <- c("purple1", "coral", "lavender")
species_names <- c("setosa", "versicolor","virginica")

plots_list <- list()

# Loop through each species dataframe
for (i in seq_along(iris_species_list)) {
  
  species_data <- ____[[i]]
  color <- ____[i]
  title <- paste(____[i], "Sepal Width Histogram")
  
  # Create the plot and store it in the list
  ____[[i]] <- plot_histogram(species_data, 
                                    x_variable = Sepal.Width, 
                                    color = color, 
                                    bins = 10,
                                    title = title,
                                    blah = theme_classic())
}

# Display the plots in a grid
grid.arrange(grobs = plots_list, ncol = 3)

```



