---
title: "Intro to Ggplot"
author: "Simone and Alys"
date: "28/06/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Ggplot

When you begin to need more functionality than Base R plotting can give you, its time to turn to ggplot. Infintely more tweakable and incredibly well documented, with ggplot you can create sleak and paper-quality plots and graphs. 

This tutorial is an introduction to ggplot with examples to get you started.

### Setting up

```{r load, message=TRUE, warning=FALSE}

basedir <- "C:\\Users\\ssteven\\OneDrive - Deakin University\\Deakin\\Coding\\coding_club_repo"
setwd(basedir)
# install.packages("ggplot2")
library(ggplot2)
# install.packages("tidyverse")
library(tidyverse)
```

## Plots 101

You need to:
1. Start by calling a ggplot using `ggplot()`
2. Specify the data using the "data" argument `ggplot(data =  )`, and the x and y variables using the aesthetics argument   `ggplot(data = , aes(x= , y = ) )`
3. Specify  what sort of plot you make ggplot to make using a second line joing to the `ggplot()` with a `+`

Lets show some examples using the cars dataset pre-loaded in R

```{r data}
# load the data
data(iris) 

# Take a look at the data
head(iris)
```

# Setting data and x y axes

For  all the ggplots with this data, the first line will look like this, with the x and y variables changing depending on what we want to plot:
```{r set-up}
ggplot(data = iris, aes(x = , y = ))
```

# Geoms (basic plot types)

Not sure what kind of plot you want, or what the geom name is? Check out
https://www.r-graph-gallery.com/

Some examples:

One categorical variable and one continuous
relationship between Sepal.Width by the species
```{r cat}
# Box Plot 
ggplot(data = iris, aes(x = Species, y = Sepal.Width)) +
  geom_boxplot()

# Voilin plot
ggplot(data = iris, aes(x = Species, y = Sepal.Width)) +
  geom_violin()

# Bar plot
ggplot(data = iris, aes(x = Species, y = Sepal.Width)) +
  geom_bar(stat = "identity")

# a dot plot
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()

# a line plot
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_line()

```

We can also add more than one geom to a plot. A particularly useful 
option is adding a regression line to a scatterplot to see trends

```{r scatter}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  geom_smooth()

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  scale_color_manual(values=c('red','#E69F00', '#56B4E9')) + 
  geom_smooth(method = lm)
```



# Data dependent geom aesthetics

### i.e. Adding more variables to your plot
You can add more information to your plot by mapping certain elements of the
geom to columns in your data, so the elements of the geom (say colour or size) 
changes corresponding to changes in the data.

These should always be specified INSIDE the aes() wrapper.


```{r vars}

# A bubble plot 

## without mapped aesthetic
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()

## with mapped aesthetic - size against continouus variable
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width,  size = Petal.Length)) +
  geom_point()

## with mapped aesthetic - size against a categorical variable

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width,  size = Species)) +
  geom_point()

## with mapped aesthetic - size against a categorical variable

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width,  shape = Species)) +
  geom_point()

# A violin plot
## no mapped aesthetic

ggplot(data = iris, aes(x = Species, y = Sepal.Width)) +
  geom_violin()

## with mapped aesthetic - fill of a categorical variable (fill refers to color inside lines)
ggplot(data = iris, aes(x = Species, y = Sepal.Width, fill  = Species)) +
  geom_violin()

##  with mapped aesthetic - Color for violin  on a categorical variable (color refers to lines)
ggplot(data = iris, aes(x = Species, y = Sepal.Width, color  = Species)) +
  geom_violin()

# A dot plot
## no mapped aesthetic

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()

##  with mapped aesthetic - color for point on a continuous variable
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width,  color = Sepal.Length)) +
  geom_point()

##  with mapped aesthetic - color for point on a categorical variable
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width,  color = Species)) +
  geom_point()


```

You can also use multiple of these together, either for the same or different variables. 
However, be sure that adding more is actually enhancing the point you are making 
with the graph, and not making it more difficult to understand.

E.g. sometimes mapping the same variable to multiple aesthetics might help,
but often it might just confuse the reader with redundant information.


```{r vars2}

# categorical variables - seems to help
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species, 
                        shape = Species)) +
  geom_point()

# continuous variables - confusing?
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length, 
                        size = Petal.Length)) +
  geom_point(alpha = 0.5)

```


### Choosing your own colours and sizes

While the *difference* between aesthetic elements is controlled by the data,
you can still control how you want those differences to be represented by specifying
colours, shapes etc.  If you don't specify these things, R will choose them for 
you.

Note that the viridis color palettes are designed to be colour-blind friendly
and can also be converted to black and white.

Load premade palettes

```{r cols}
# install.packages("RColorBrewer")
# install.packages("viridisLite")
library(RColorBrewer)
library(viridisLite)

# To see the color palettes from RColorBrewer
display.brewer.all()
```

When you need to "fill" for a categorical variable:
```{r cols_fill_cat}
# categorical variables - colour
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point()

# categorical variables - size
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, size = Species)) +
  geom_point() +
  scale_size_manual(values = c(1, 3, 6))

# Colour by name or HEX
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  scale_color_manual(values = c("darkgreen", "purple", "#4842f5"))

# Colour from a preloaded palette
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  scale_color_brewer(palette="Set2")

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  scale_color_viridis_d() # d for discrete

# Grey
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  scale_color_grey()

```


When you need to "fill" for a continuous variable:
```{r cols_fill_cont}
# categorical variables
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length)) +
  geom_point()

# Colour by name or HEX
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length)) +
  geom_point() +
  scale_color_gradient(low="blue", high="red")

# Colour from a preloaded palette
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length)) +
  geom_point() +
  scale_color_gradientn(colours = rainbow(5))

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length)) +
  geom_point() +
  scale_color_distiller(palette = "Greens") # the palette is chosen from display.brewer.all()

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length)) +
  geom_point() +
  scale_color_viridis_c() # c for continuous

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length)) +
  geom_point() +
  scale_color_viridis_c(option = "plasma") # choosing a palette. try "magma", or "inferno"

```

When you need to "color":
```{r cols_color}
# categorical variables
ggplot(data = iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_boxplot()

# Colour by name or HEX
ggplot(data = iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_boxplot() +
  scale_fill_manual(values = c("green", "pink", "#4842f5"))

# Colour from a preloaded palette
ggplot(data = iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_boxplot() +
  scale_fill_brewer(palette="Set2")

ggplot(data = iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_boxplot() +
  scale_fill_viridis_d()

# Grey
ggplot(data = iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_boxplot() +
  scale_fill_grey()

```

# Data independent geom aesthetics

You can also set these parameters to be constant and without being associated to 
a variable by not putting them into the aes() call, and instead specifying what 
you want it to look like.

```{r}
# Colour by a name and shape
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(shape = 17, color  = "darkgreen")

# Colour by a HEX and size
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(alpha = 0.5, color = "#4842f5", size = 0.8)

```

# Data independent, non-geom aesthetics

These are modifications to the architecture of your plot that are not specific to
the data OR the geom.  These can include titles, font size and type, tick marks, 
background and so on. 


Add a title

```{r title}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()+
  labs(title = "Sepal length by width")
```

Change the x and y labels
```{r labs}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()+
  labs(title = "Sepal length by width", x = "Sepal length (cm)", y="Sepal width (cm)")
```

Box plot with flipped coordinates
```{r box flip}
ggplot(data = iris, aes(x = Species, y = Sepal.Width)) +
  geom_boxplot() +
  coord_flip()
  
```

# Themes

These are changes we make to the plot architecture - axes, backgrounds, fonts,
grids, legends etc.

We can tweak individual elements inside the 'theme' call.


Text size and angled
```{r text modify}
ggplot(data = iris, aes(x = Species, y = Sepal.Width)) +
  geom_boxplot() +
  theme_classic() +
  theme(text = element_text(size = 15), axis.text.x = element_text(angle = 45, hjust = 1) ) 
  
```


Legend position
```{r legend}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  theme(legend.position = "bottom") # try "top" or "none" to remove the legend altogether
```

OR, we can use pre-made themes to save time and keep our plots consistent

Change the theme to black and white
```{r box theme}
ggplot(data = iris, aes(x = Species, y = Sepal.Width)) +
  geom_boxplot() +
  theme_bw() # try theme_light, theme_dark, theme_classic,

?theme_bw()
  
```

###  Theme assist (life changer!)

The most annoying thing about making plots is the trial and error process of
writing the code, looking at the plot, amending the code, looking again etc etc. 
Theme assist
can speed up the 'theme' part of code writing.

Write your basic plot code, specifying data, geom, axes and label but don't
worry about adding any code for the theme.

After installing the package, highlight the code then click on 'Addins' at the 
top of R Studio and 
select ggthemeassist

``` {r theme assist}

# install.packages("ggThemeAssist") # install like a package
library(ggThemeAssist)

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() 

```

OR we can get really clever and have the best of both worlds by specifying our 
own pre-set theme. That way if you want to tweak one thing, it will automatically
change it in all the plots you've applied it to, instead of having to go and
adjust say, text size, for each plot code you've written.

``` {r personal theme}

# set some of the elements that apply to multiple arguments inside the theme

background.colour <- "pink"
text.colour <- "black"
text.font <- "Calibri"
line.width <- 0.6

# copy paste the theme info from the theme assist

best_theme_ever <- theme(plot.subtitle = element_text(size = 15,
    colour = "cadetblue4"), legend.title = element_text(size = 14),
    legend.key = element_rect(fill = NA),
    legend.background = element_rect(fill = NA),
    legend.position = "bottom", legend.direction = "horizontal")

ggplot(data = iris, aes(x = Species, y = Sepal.Width)) +
  geom_boxplot() +
  best_theme_ever
 

```

# Getting fancy

### Facets

Faceting into column - this is also mapped to data but is not specific to the
geom, you can apply it to most plot types provided your data is properly formatted

```{r facet1}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() +
  theme_bw() +
  facet_wrap(~Species, ncol = 3) # try ncol = 1, or instead of ncol write nrow

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() +
  theme_bw() +
  facet_wrap(~Species, nrow = 3) 
  
```

Faceting into rows with your own labels
```{r facet2}
  
  ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() +
  theme_bw() +
  facet_wrap(~Species, ncol = 1,
             labeller = labeller(Species = c("setosa" = "Iris setosa",
                                             "versicolor" = "Iris versicolor",
                                             "virginica" = "Iris virginica")))
  
```


Breaks and grid lines
```{r breaks}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() +
  scale_x_continuous(limits = c(4, 8),
                     breaks = seq(4, 8, 1), # sequence from 0 to 8 by 1. Equivalent to 0:8
                     minor_breaks = seq(0, 8, 0.2)
                     )
  
  
```

### Annotations

For adding text to plots. There is also an option to add geom_text, but this
is more computationally expensive, so it's usually better to use annotate
when you can.


```{r annotate}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  annotate("text", x = 4.5, y = 4, label = "Group 1", size = 7) +
  annotate("text", x = 6.5, y = 3.6, label = "Group 2", color = "blue") + 
  annotate("text", x = 5.5, y = 2, label = "Group 3") 
  
# Bar plot
ggplot(data = iris, aes(x = Species, y = Sepal.Length)) +
  geom_bar(stat = "identity") +
    annotate("text", x = 1, y = 270, label = "*", size = 20) 
  
```

Draw arrows
```{r draw_lines}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  annotate("segment", x = 4, xend = 4, y = 2.8, yend = 3.5)

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  annotate("text", x = 4.5, y = 4.2, label = "Important segment") +
  annotate("segment", x = 4.1, xend = 5, y = 4, yend = 4,
           arrow = arrow(ends = "both", angle = 90, length = unit(.2,"cm")))

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  annotate("text", x = 7, y = 4.5, label = "Outliers", size = 7) + # try size = 3
  annotate("segment", x = 7.4, xend = 7.8, y = 4.5, yend = 3.9, # try chaning these values
           size = 1.5, # try 0.5 or 2
           color = "darkblue", # try "green" or "purple"
           arrow = arrow())
  
```

### Building  a plot up step by step

```{r building up}
p <- ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()

p

p + theme_bw()
p2 <- p + theme_bw()

p2 + labs(title = "Add title", subtitle = "And a sub-title", 
          y = "Sepal width (cm)", x = "Sepal length (cm)")
  
```


### Save your plot

No more click and points exporting or screenshots.  

```{r save}

my_plot <- ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
           geom_point()

help(ggsave) # if you can't remember what order to specify the arguments check here

ggsave("my_plot", 
       my_plot, 
       device = "jpeg",
       dpi = 300 ) # Resolution - for publication needs to be 300 or above

```


### Integration with other packages
### Building  a plot up step by step
```{r ggplotly}
# install.packages("plotly")
library(plotly)

p <- ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()

ggplotly(p)
  
```
Change the text for the mouse hover
```{r ggplotly2}
p <- ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, text = paste0("Sepal length: ", Sepal.Length, "cm <br>", "Sepal width: ", Sepal.Width, "cm <br>", "Petal length: ", Petal.Length, "cm <br>", "Petal width: ", Petal.Width, "cm <br>", "Species: ", Species) )) +
  geom_point()

ggplotly(p, tooltip = "text")
  
```

### Cropping/zooming a plot

When wanted to crop a plot, use coord_cartesian()rather than scaling the axis 
because any statistical analysis done on the plot (eg adding regression line) 
changes. This is because the stats are based on the data in the plot. Using crop 
cuts out data from the data frame so the stat analysis changes. 
Using coord_cartesian only changes what you are seeing

```{r coord cartesian}

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_smooth() 

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_smooth() +
  coord_cartesian(xlim = c(3, 6)) # Only show us the section of the line with sepal length between 3 and 6

  
```

### Arranging multiple plots

Perhaps you don't want to use facet_wrap for some reason (for example, if
you want different legends on each plot), you can use other functions to
arrange your plots.

```{r multiple plots}

# install.packages("cowplot")
library(cowplot)

setodf <- iris %>% filter(Species == "setosa")
versdf <- iris %>% filter(Species == "versicolor")
virgdf <- iris %>% filter(Species == "virginica")

setosa_plot <- ggplot(data = setodf, aes(x = Sepal.Length, y = Sepal.Width,
                                         color = Petal.Width)) +
  geom_point() +
  scale_color_viridis_c(option = "inferno") # c for continuous

virginica_plot <- ggplot(data = virgdf, aes(x = Sepal.Length, y = Sepal.Width,
                                           color = Petal.Width)) +
  geom_point() +
  scale_color_viridis_c() # c for continuous

versicolor_plot <- ggplot(data = versdf, aes(x = Sepal.Length, y = Sepal.Width,
                                           color = Petal.Width)) +
  geom_point()  +
  scale_color_viridis_c(option = "magma") # c for continuous



plot_grid(setosa_plot, virginica_plot,versicolor_plot, align = "v", 
          nrow = 3, rel_heights = c(1/4, 1/4, 1/4))

```

### Align mismatched axes and legends - TODO

``` {r grid arrange}

# install.packages("gridExtra")
library(gridExtra)

sepal_plot <- ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width,
                                         color = Species)) +
  geom_point() +
  scale_color_viridis_d(option = "inferno") # d for discrete

petal_plot <- ggplot(data = iris, aes(x = Petal.Length, y = Petal.Width)) +
  geom_point() +
  scale_color_viridis_d() # d for discrete

sepal_plot

petal_plot

grid.arrange(sepal_plot, petal_plot)

plot_grid(sepal_plot, petal_plot, align = "V", 
          nrow = 2, rel_widths = c(1/4, 1/4))

# install.packages("gtable")

library(gtable)

g2 <- ggplotGrob(sepal_plot)
g3 <- ggplotGrob(petal_plot)
g <- rbind(g2, g3, size = "first")
g$widths <- unit.pmax(g2$widths, g3$widths)
grid.newpage()
grid.draw(g)

```
Adding two datasets to the same plot - this is particularly useful for maps

```{r maps}

#install.packages("sf")
library(sf)

# Load coastline polygons

coast_nz_shp <- st_read("nz-coastlines-topo-150k.shp", quiet = TRUE)

# Load port locations

all_port_points <- st_read("wld_trs_ports_wfp.shp", quiet = TRUE)

# Filter to only the points in new zealand

ports_nz_points <- all_port_points %>% 
                   st_as_sf() %>% 
  filter(country == "New Zealand") %>% 

# Re-project the points to NZ GD 2000 CRS

ports_nz_points <- st_transform(ports_nz_points, 4167)

map <- ggplot() + 
       geom_polygon(data = shp, 
       aes(x = long, y = lat, group = group), colour = "black", fill = NA) +
       geom_point(data = ports_nz_points, )
```

