---
title: "Intro to Ggplot"
author: "Simone and Alys"
date: "28/06/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Ggplot

When you begin to need more functionality than Base R plotting can give you, its time to turn to ggplot. Infintely more tweakable and incredibly well documented, with ggplot you can create sleak and paper-quality plots and graphs. 

This tutorial is an introduction to ggplot with examples to get you started.

### Setting up

```{r load, message=TRUE, warning=FALSE}
install.packages("ggplot2")
library(ggplot2)
```

## Plots 101

You need to:
1. Start by calling a ggplot using `ggplot()`
2. Specify the data using the "data" argument `ggplot(data =  )`, and the x and y variables using the aesthetics argument   `ggplot(data = , aes(x= , y = ) )`
3. Specify  what sort of plot you make ggplot to make using a second line joing to the `ggplot()` with a `+`

Lets show some examples using the cars dataset pre-loaded in R

```{r data, echo=FALSE}
# load the data
data(iris) 

# Take a look at the data
head(iris)
```
For  all the ggplots with this data, the first line will look like this, with the x and y variables changing depending on what we want to plot:
```{r set-up, echo=FALSE}
ggplot(data = iris, aes(x = , y = ))
```

### Basic sorts of plots
Scatter plot
relationship between Sepal.Length and Sepal.Width
```{r scatter, echo=FALSE}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()
```

Box plot
relationship between Sepal.Width by the species
```{r box, echo=FALSE}
ggplot(data = iris, aes(x = Species, y = Sepal.Width)) +
  geom_boxplot()
```


bar plot
geom_bar

line
geom_line
geom_trace  

### Not so basic plot structure
violin


### Basic modifcations
Add a title
```{r scatter title, echo=FALSE}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()+
  labs(title = "Sepal length by width")
```

Change the x and y labels
```{r scatter labs, echo=FALSE}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()+
  labs(title = "Sepal length by width", x = "Sepal length (cm)", y="Sepal width (cm)")
```



shape by group
```{r scatter, echo=FALSE}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point()
```

Lengend position
```{r scatter, echo=FALSE}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  theme(legend.position = "bottom")
```


colour, shape and size can change by group
```{r scatter, echo=FALSE, warning=FALSE}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, shape = Species, color = Species, size = Species)) +
  geom_point()
```

colour by group with your choice of colours
```{r scatter, echo=FALSE}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  scale_color_manual(values=c('red','#E69F00', '#56B4E9'))
```
More on colour here
scale_fill_manual -  use a bar plot to show this




Add a regression line to see trends
```{r scatter, echo=FALSE}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  scale_color_manual(values=c('red','#E69F00', '#56B4E9')) + 
  geom_smooth(method=lm)
```

### Customise further
add other arguments for customisation

Box plot with flipped coordinates
```{r box flip, echo=FALSE}
ggplot(data = iris, aes(x = Species, y = Sepal.Width)) +
  geom_boxplot() +
  coord_flip()
  
```


Change the theme
```{r box theme, echo=FALSE}
ggplot(data = iris, aes(x = Species, y = Sepal.Width)) +
  geom_boxplot() +
  theme_bw()
  
```

Text size and angled
```{r text modify, echo=FALSE}
ggplot(data = iris, aes(x = Species, y = Sepal.Width)) +
  geom_boxplot() +
  theme_classic() +
  theme(text = element_text(size = 15), axis.text.x = element_text(angle = 45, hjust = 1) ) 
  
```


Facetting
```{r facet1, echo=FALSE}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() +
  theme_bw() +
  facet_wrap(~Species, ncol=3)
  
```

Facetting with your own labels
```{r facet2, echo=FALSE}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() +
  theme_bw() +
  facet_wrap(~Species, ncol=1, labeller = labeller(Species =
                                                              c("setosa" = "Iris setosa",
                                                                "versicolor" = "Iris versicolor",
                                                                "virginica" = "Iris virginica")))
  
```


### Building  a plot up step by step
```{r building up, echo=FALSE}
p <- ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()

p

p + theme_bw()
p2 <- p + theme_bw()

p2 + labs(main="Add title")
  
```

### Integration with other packages
### Building  a plot up step by step
```{r ggplotly, echo=FALSE}
install.packages("plotly")
library(plotly)

p <- ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()

plotly::ggplotly(p)
  
```
