---
title: "Intro to species information packages"
author: "Simone Stevenson"
date: "28/07/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Species packages

There are multiple R packages available for accessing biodiversity data, including:
Taxonomic information, ecological traits, distribution, occurrence points, conservation
status etc. These become particularly useful when you need data for a list of
different species, although you cannot guarantee the same data will be
available for each species.  Therefore the code below may not work for every 
species you try it on.

With the exception of Galah, the packages below were all developed by someone
called Scott Chamberlain.  Therefore they work in very similar ways so if you 
learn one, you can probably figure the others out too. 

NOTE: Nearly all databases return objects in the form of lists that you then 
need to subset to get the information you need. Getting familiar with lists and 
how to subset them is probably a good idea if you want to use any of these 
packages

### Setting up

```{r packages, echo=FALSE, message = FALSE, warning = TRUE}

# set directory

basedir <- "C:/Users/ssimone/OneDrive - Deakin University/Coding_club/231124_simone_spp_data"
setwd(basedir)

# load packages

## for data wrangling + ggplot
# install.packages("tidyverse")
library(tidyverse)

## for getting species info
# install.packages("taxize")
library(taxize)
# install.packages("rgbif")
library(rgbif)
# install.packages("galah")
library(galah)
# install.packages("rredlist")
library(rredlist)
# install.packages("rfishbase")
# library(rfishbase)

# for mapping
# install.packages("sf")
library(sf)
# install.packages("rnaturalearth")
library(rnaturalearth)
# install.packages("rnaturalearthdata")
library(rnaturalearthdata)

```

# Species information inputs

```{r species, message=TRUE, warning=FALSE}

# Load your species name here. Some packages will also take a list of species.
# For others you would need to use an apply function or a loop.

spp_name <- "Carcharodon carcharias"
spp_name

# If you're not sure how to pronounce Carcharodon carcharius like I wasn't,
# youtube is your friend https://www.youtube.com/watch?v=yqwonwErTNc&ab_channel=OCEARCH

```

# taxize

Taxize is a package that lets you check for species taxonomic information and
whether your species has synonyms - that is, is it/has it been known by multiple
different scientific names. If it has, you may want to search all known synonyms
when looking to download data from species databases.

Citation:
Chamberlain, S. A., & Szöcs, E. (2013). taxize: taxonomic search and retrieval in R. F1000Research, 2.

```{r taxized}

# Get the unique taxonomic serial number for your species

spp_id <- get_tsn(spp_name)

# Get a list of its synonyms
synonyms_list <- synonyms(spp_id, db = "itis")

# Turn into a nice dataframe

spp_synonyms_df <- do.call(rbind, synonyms_list)
spp_synonyms_df

# Make a vector of the various synonyms your species is known by
spp_synonyms <- c(spp_name, spp_synonyms_df$syn_name)
spp_synonyms

# Get taxonomic classification information for your species
taxonomic_info_list <- classification(spp_name, db = "itis")

# Turn into a nice dataframe
taxonomic_info_df <- do.call(rbind, taxonomic_info_list)

taxonomic_info_df

```

#rredlist

The rredlist package talks to the IUCN Red List website and allows you to find
information on current and past conservation status of species, taxonomic info,
the reason they were listed, and their key threats. 

Global in scope.

Does not include spatial information, and you need to have obtained an access
token from the red list website and stored it. I can help you do this if you get
stuck.

IUCN: Request an API key at http://apiv3.iucnredlist.org/api/v3/token. Note it 
can take a few days to receive.

Post showing how to save the key:https://stackoverflow.com/questions/45185852/accessing-an-api-with-r

"You should also put your API key in something like an environment variable so 
it never ends up in scripts or GitHub repo commits. Use ~/.Renviron (make a 
single line entry for IUCN_REDLIST_KEY=your_key and then restart R)."

Citation:
Chamberlain S. 2018. rredlist: 'IUCN' Red List Client

```{r rredlist}
# Command to add your token (will open your .environ file, store as 
# IUCN_REDLIST_KEY = "your token")

usethis::edit_r_environ()

red_list_synonyms <- rl_synonyms(spp_name)[["result"]]
red_list_synonyms

spp_red_list_info <- rl_search(spp_name)[[2]]

spp_conservation_status <- spp_red_list_info %>% 
                           select(category)

spp_conservation_status

spp_red_list_history <- rl_history(spp_name)[[2]] %>% 
                        arrange(year)

spp_red_list_history

spp_red_list_threats <- rl_threats(spp_name)[[2]] 

spp_red_list_threats
  
```

# rgbif

GBIF (Global Biodiversity Information Facility) is an online database of species
occurrence records. It is global in scope, and has been assembled from a vast
number of field observations, of varying age and quality.

Global in scope and contains occurrences from multiple taxa.

Recommended to do some quality checking of your outputs.  If high accuracy is 
important (eg for SDMs or occupancy modeling or abundance), you may be better 
off searching for a more carefully curated database specific to your species 
and/or location.

Don't think it requires an API to access.

Citation: Chamberlain S, Barve V, Mcglinn D, Oldoni D, Desmet P, Geffert L, 
Ram K. 2020. rgbif: Interface to the Global Biodiversity Information Facility API

```{r rgbif}

# Get an identifying key for your species

occ_key <- name_suggest(spp_name)

# extract it from the output list

keys <- occ_key[["data"]]$key[1]

# Check it is a single number (or multiple numbers if searching multiple spp)
keys

# Download data from GBIF database. Note 'limit' sets the number of observations
# you want to download. A lower number will be quicker but incomplete. Larger
# numbers are obviously more complete but take longer. I think it downloads most
# recent first, then works backwards.

# You might want to check how many observations there are in total first so
# you know how incomplete your set will be if you set the limit lower

spp_occ_count <- occ_count(keys)
spp_occ_count

# Get your occurrences
gbif_data_list <- occ_search(keys, limit = 200) 

# Get taxonomic info (less detailed than the taxize version)
names(gbif_data_list)

gbif_taxonomic_info <- gbif_data_list[["hierarchy"]][[1]]

gbif_taxonomic_info

# Get the observations and their associated attributes

gbif_data <- gbif_data_list[["data"]]

# Filter whatever columns you care about

gbif_occurrences <- gbif_data_list[["data"]] %>%
        select(genericName, specificEpithet, 
               decimalLatitude, 
               decimalLongitude,
               lifeStage, iucnRedListCategory,year) 

# Have a look
head(gbif_occurrences)

# Generate citations (these are for the individual field datasets)
rgbif_citations <- gbif_citation(gbif_data_list)
rgbif_citations[[1]]

```

# galah

Galah is the package for accessing the Atlas of Living Australia, another
online database of species occurrence records. I *think* it is limited to species
found in Australia but I'm not sure. It can give you occurrence points outside
Australia too though.

Doesn't require an API but you do need to have registered an account on the
ALA website, and then configure below with your registered email address.

Probably the easiest package to use and I think at least some of the species data
has undergone decent quality control but still always good to check outputs.

Citation:
Stevenson M, Westgate M, Newman P (2021). galah: Atlas of Living Australia (ALA) Data and Resources in R. R package version 1.2.0, https://CRAN.R-project.org/package=galah.

```{r galah}

# Configure your email

ala_config(email = "ssteven@deakin.edu.au")

# Get the species info you need to access the database
ala_spp <- select_taxa(spp_name)

# Get occurrence data for your species. You can filter by various geographic
# subunits including country and stateProvince. Adding the 'mint_doi' argument
# means you can also generate a citation for your data, recommended.

occ <- ala_occurrences(ala_spp,
                       select_filters(country = "Australia"),
                       mint_doi = TRUE)

# if you want to save
# write.csv(occ, file.path(basedir, "spp_occurrences.csv"))

# Generate a citation
citation <- ala_citation(occ)
citation

```

# plotting occurrence data

You can plot the data which is pretty fun. Some very basic code for converting
the occurrence data into spatial objects and making a basic map below. This is
good for doing a quick check that your data looks sensible. Will need better
code or export the shapefiles for GIS to make a publication ready map though.

```{r plot gbif}

# Remove incomplete lines
gbif_occurrences_complete <- gbif_occurrences[!with(gbif_occurrences,is.na(decimalLatitude)&is.na(decimalLatitude)),]

# Convert to an sf object
gbif_coords_shp <- st_as_sf(gbif_occurrences_complete, 
                            coords = c("decimalLongitude",
                                       "decimalLatitude"),
                       crs = 4326)

# Save your shapefile
st_write(gbif_coords_shp, "gbif_occcurrences.shp")

# Download a polygon using the rnatural earth package

world_shp <- ne_countries(scale = "medium", returnclass = "sf")

# Plot the points on the polygon
ggplot() +
    geom_sf(data = world_shp) +
    geom_sf(data = gbif_coords_shp, aes(col = lifeStage)) + # colour by lifestage
    scale_color_viridis_d(option = "viridis", na.value = "hot pink")


```


```{r plot ala}

# ALA data 
# Remove incomplete lines
occ <- occ[!with(occ,is.na(decimalLatitude)& is.na(decimalLatitude)),]

# convert to sf object
occ_coords_shp <- st_as_sf(occ, coords = c("decimalLongitude",
                                       "decimalLatitude"),
                       crs = 4326)

# Remove unneccessary attributes if you like (good if you have lots of observations)
occ_coords_shp <- occ_coords_shp %>%
                  dplyr::select(scientificName, geometry)

# Download a polygon, this time lets just get australia

aus_shp <- ne_countries(scale = "medium", returnclass = "sf", country = "Australia")

# Plot the points over the polygon
ggplot() +
    geom_sf(data = aus_shp) +
    geom_sf(data = occ_coords_shp, col = "hot pink")


```


# fishbase

I ran out of time to figure out the fishbase package!  The advantage of using 
fishbase is that it appears to have a lot of additional attributes you won't
get from the other databases, including species trait information and ecosystem
type.  We can come back to it another day.

Citation:
Boettiger, C., Chamberlain, S., Lang, D. T., Wainwright, P., & 
Boettiger, M. C. (2021). Package ‘rfishbase’.

```{r fishbase}

# no code sorry :(

```



